# Plotting elegant graphs

Once we have seen what ggplot is capable of, it is now time to delve into graphics and to start understanding better the philosophy behind ggplot (and by extension of the whole tidyverse).

```{r}
#| echo: false
#| warning: false
library(tidyverse)
```

## The Palmer Penguin data set

We are going to use the Palmer Penguin data set, which is a data set containing the size measurements for adult foraging penguins near Palmer Station, Antarctica. This data set is contained in an external package that can be installed, as usual, using the command `install.packages("palmerpenguins")`. To load the package, we can now use the command

```{r}
library(palmerpenguins)
```

Before we can start plotting graphs with this data set, let's start by checking the data set:

```{r}
penguins
glimpse(penguins)
```

This is a small data set, with only 344 data. Also observe that some factor variables (`island`, `species`, and `sex`) and that the other variables are numeric. We can also see that there are some NAs in the data set.

Before starting we also need to get a feeling of how many levels the factors have. Let's just go for the simplest way which is using the `summary()` function to see this:

```{r}
summary(penguins)
```

So overall there are three species: Adelie, Chinstrap and Gentoo, and thee islands: Biscoe, Dream and Torgersen. The sex of the penguins is balanced, and there are only 11 NAs.

## Scatter plots

The first type of graph we want to learn how to do is a scatter diagram. We also want to learn how to add a regression line (line of best fit) to a scatter diagram and learn a few parameters that we may wish to change when doing scatter diagrams.

Using the penguin data set, let's start by sketching a scatter diagram of the body mass(in g) against the flipper length (in mm). The basic code is the following:

```{r}
penguins |> 
    ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point()
```

Notice that ggplot returns a warning that 2 rows have been removed because they contain NAs. To avoid this warning, we could remove the NAs from the two variables, using the function `drop_na()` from **tidyr**. We also want to distinguish the three species in the graph, so we want to map the `species` variable to another geometrical feature of the graph. For example, we could map the `species` to the shape of the point, like so:

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |> 
    ggplot(aes(x = flipper_length_mm, y = body_mass_g, shape = species)) +
    geom_point()
```

However this is not fully clear and some of the points are overlapping, making the diagram even less clear. How about mapping the `species` to different colours? Like so:

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g, colour = species)) +
    geom_point()
```

This is way better, but can we do even better? How about both?

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |> 
    ggplot(aes(x = flipper_length_mm, y = body_mass_g,
               shape = species, colour = species)) +
    geom_point()
```

Let's now add a line of best fit. This is done using the `geom_smooth()` function, to which we need to pass the method (in our case a **linear model**, `lm`, with the formula `y ~ x` – `y` is the response variable and `x` is the explanatory variable). Precisely

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g,
               shape = species, colour = species)) +
    geom_point() +
    geom_smooth(method = "lm", formula = "y ~ x")
```

This isn't quite what we had in mind though. Why? Because passing the aesthetics `shape = species` and `colour = species` to the `ggplot()` function implies that these mappings are passed globally to `geom_point()` (as we wanted) and to `geom_smooth()` that interprets them as our request to draw a regression line for each group. To have one single regression line, we need to rearrange our code a little, passing the `shape = species` and `colour = species` aesthetic mappings to `geom_point()` so that these are interpreted locally, like so:

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species, colour = species)) +
    geom_smooth(method = "lm", formula = "y ~ x")
```

I also find the error bar around the line of best fit incredibly ugly, and often not particularly informative. To get rid of it, it suffices to pass the parameter `se = FALSE` to `geom_smooth()`, like so:

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species, colour = species)) +
    geom_smooth(method = "lm", formula = "y ~ x", se = FALSE)
```

Let's also change the colour of these points. To do this, we can use one of the `scale_colour_` functions. Precisely, we are going to use the `scale_colour_brewer` which allows to access to several colour palettes. Some people use `scale_colour_viridis_d` but I find the viridis palettes to be horribly ugly, so I always steer clear from this one.

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species, colour = species)) +
    geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
    scale_colour_brewer(palette = "Set2")
```

Now we wish to re-scale the y-axis to make it more accurate. This can be done using the `scale_y_continuous`. To use this we need to pass to the parameter `breaks` a sequence going from the minimal value to the maximum value and the number of breaks in this sequence.

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species, colour = species)) +
    geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
    scale_colour_brewer(palette = "Set2") +
    scale_y_continuous(breaks = seq(1000, 8000, 500))
```

Finally, let's change the labels on the axes. This can be done with the function `labs()`. Let's also add a title and a subtitle.

```{r}
penguins |> 
    drop_na(flipper_length_mm, body_mass_g) |>
    ggplot(aes(x = flipper_length_mm, y = body_mass_g)) +
    geom_point(aes(shape = species, colour = species)) +
    geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
    scale_colour_brewer(palette = "Set2") +
    scale_y_continuous(breaks = seq(1000, 8000, 500)) +
    labs(
        title = "Body mass and flipper length",
        subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
        x = "Flipper Length (mm)", y = "Body Mass (g)", 
        colour = "Species", shape = "Species"
    )
```

This is a rather interesting scatter diagram, much more informative than the first attempt. This graph could now be saved and used into a presentation, or into an article, etc.

## Bar charts

Let us now draw a bar chart. For example, let's draw the frequency distribution of the species in the Penguin data set. The basic code is the following:

```{r}
penguins |> 
    ggplot(aes(x = species)) +
    geom_bar()
```

In bar plots as such, where the categorical variable has non-ordered levels, it is often preferable to reorder the bars based on their frequencies. We can do this using the **forcats** function `fct_infreq()`, like so:

```{r}
penguins |> 
    ggplot(aes(x = fct_infreq(species))) +
    geom_bar()
```

Let's now make some changes. First of all, let's add some transparency. This is done using the parameter `alpha` which is a number that goes from 0 (total transparency) to 1 (no transparency). For example, let's set `alpha = 0.7`.

```{r}
penguins |> 
    ggplot(aes(x = fct_infreq(species))) +
    geom_bar(alpha = 0.7)
```

We also wish to change the colour of the bars. We don't want to map a variable to the colour, but we only wish to change all colours. To do so, it suffices to change the `fill` parameter. R has a number of colours already defined for you to use. A full list can be be seen by typing the command `colors()` in the console. Let's go for "midnightblue" which is Julia Silge's favourite (and I don't dislike it either):

```{r}
penguins |> 
    ggplot(aes(x = fct_infreq(species))) +
    geom_bar(alpha = 0.7, fill = "midnightblue")
```

Finally, we want to change the labels and add a title to the graph. The x-axis is kind of obvious, so we don't really need a label at all. We can pass `x = NULL` to tell ggplot to skip the label.

```{r}
penguins |> 
    ggplot(aes(x = fct_infreq(species))) +
    geom_bar(fill = "midnightblue", alpha = 0.7) +
    labs(
        title = "Distribution of species of the Palmer Penguin data set",
        x = NULL, y = "Frequency"
    )
```

Note that `geom_bar()` only requires you to assign the x-axis because by default it calculates the frequencies of the values on the x-axis.

## Histograms

Let's now plot a histogram for the body mass of the penguins. The basic code is the following:

```{r}
penguins |> 
    ggplot(aes(x = body_mass_g)) +
    geom_histogram()
```

Again we may wish to remove the missing values ourselves, to avoid the warning, and we may wish to change the bin width to try make a few attempts and see which one is the most informative value. Let's try with two extremes:

```{r}
penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_histogram(binwidth = 20) 

penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_histogram(binwidth = 2000)
```

When, like in this case, it turns out that choosing a bin width may be complicated, it is quicker to opt for a fixed number of bins, like so:

```{r}
penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_histogram(bins = 20)
```

Again, let's change transparency and colour and let's change the labels on the axes:

```{r}
penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_histogram(bins = 20, fill = "midnightblue", alpha = 0.7) +
    labs(x = "Body Mass (g)", y = "Frequency")
```

## Density diagrams

A similar type of diagram is the density diagram.

```{r}
penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_density(colour = "midnightblue", fill = "midnightblue", alpha = 0.6)
```

If we want to study the relationship between body mass and the species, we may wish to draw the diagram by species. This can be done passing variable `species` to the `colour` and the `fill` parameter. This however must be done inside the `aes()` function because it is an aesthetic mapping.

```{r}
penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_density(aes(colour = species, fill = species), alpha = 0.4)
```

I find the numbers on the y-axis to be really ugly and would rather having normal numbers there. There is an add-on packages in the **tidyverse** that allows us to deal with formats (such as percentages, currencies, etc). This is called `scales`. In this case, let's use the function `scales::number` to show the number in the ordinary way. Let's also change the labels on the axes.

```{r}
penguins |>
    drop_na(body_mass_g) |> 
    ggplot(aes(x = body_mass_g)) +
    geom_density(aes(colour = species, fill = species), alpha = 0.4) +
    scale_y_continuous(label = scales::comma) +
    labs(x = "Body Mass (g)", y = "Density", 
         colour = "Species", fill = "Species")
```

This is now a beautiful graph, ready to be exported and used. Notice how clearly the graph describes that Adelie and Chinstrap tend to have the same average body mass, whereas the Gentoo penguins tend in average to have a much higher body mass.

## Box plots

When it comes to comparing numerical quantities, box plots can be really useful. The simplest way to build a box plot is the following:

```{r}
penguins |> 
    drop_na(body_mass_g) |> 
    ggplot(aes(x = species, y = body_mass_g)) +
    geom_boxplot()
```

Let's make the diagram prettier by changing the colours of the boxes and by changing the label on the axes.

```{r}
penguins |> 
    drop_na(body_mass_g) |> 
    ggplot(aes(x = species, y = body_mass_g,
               colour = species, fill = species)) +
    geom_boxplot(alpha = 0.3, show.legend = FALSE) +
    labs(x = "Species", y = "Body Mass (g)")
```

Again, look how this graph gives the same information we had seen before: Adelie and Chinstrap have virtually the same body mass, with more variation in the Adelie species, and Gentoo penguins have a much higher body mass. See how only one Chinstrap outlier actually falls in the Gentoo interquartile range.
